# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
The data for this report can be found at https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip.  

After downloading the data set into a csv file named, "activity.csv", it is read into R to analyze the numbers.

``` {r}
activity <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

Here is a histogram showing the frequencies of the individual's number of steps per day. 

``` {r}
library(dplyr)
library(ggplot2)
steps_day <- activity %>% group_by(date) %>% summarise(T_steps = sum(steps,na.rm = T))
library(ggplot2)
ggplot(steps_day,aes(T_steps))+
    geom_histogram(bins = 20)
```

The **mean** and **median** total number of steps taken per day can be calculated by looking at the daily totals tallied in the *daily* variable above.

``` {r}
stepsMean <- mean(steps_day$T_steps)
stepsMedian <- median(steps_day$T_steps)
stepsMean
stepsMedian
```
On days when data was collected, the individual in this case took an average of `r stepsMean` steps per day with a median value of `r stepsMedian` steps.

## What is the average daily activity pattern?
The following plot displays the individual's average number of steps by 5-minute interval over the entire two months.  Entries of "NA" are once again omitted.  

``` {r}
steps_interval <- activity %>% group_by(interval) %>% summarise(T_steps = mean(steps,na.rm = T))
library(ggplot2)
ggplot(steps_interval,aes(x=interval,y=Average_steps))+
    geom_line()
```

The interval with the highest average number of steps and its value are:

``` {r}
steps_interval[which.max(steps_interval$Average_steps),]
```

## Imputing missing values

This data set has several missing values, denoted by "NA."  The number of such instances can be found easily by:  

Let's count how many missing values exist for each variable 

``` {r}
colSums(is.na(activity))
```
There are 2304 cases with NA's in the feature *steps*.

To get a complete set of data, "NA" entries will be replaced with the average steps taken value across of each interval data set.  
This will be done in a new data set, named *new_data*.  A histogram of steps taken each day with this new data set, as well as its **mean** and **median** are all displayed below. 

``` {r}
steps_interval <- activity %>% group_by(interval) %>% summarise(Average_steps = mean(steps,na.rm=T))
new_data <- activity %>% inner_join(steps_interval) %>% 
    mutate(New_steps=ifelse(is.na(steps),Average_steps,steps)) %>% 
    select(-steps,-Average_steps)
    
ggplot(new_data,aes(New_steps))+
    geom_histogram(bin=30)

stepsCompleteMean <- mean(new_data$New_steps)
stepsCompleteMedian <- median(new_data$New_steps)
stepsCompleteMean
stepsCompleteMedian
```

The **mean** and **median** are virtually the same as they were for the data set that omitted "NA" values.  This is because the "NA" values were replaced by the average value of each interval, which of course would not alter the overall average.  The one noticeable change occurs in the histogram, where the most frequent range of daily steps taken increased in frequency.  This is due to the days with "NA" values being assigned the mean, pushing that range higher.  

## Are there differences in activity patterns between weekdays and weekends?
``` {r}

library(ggplot2)
new_data <- new_data %>% 
    mutate(date = as.Date(date,format="%Y-%m-%d")) %>% 
    mutate(w_day = wday(date, week_start = getOption("lubridate.week.start", 1))) %>% 
    mutate(dayType = as.factor(ifelse(w_day==6,"Weekend",ifelse(w_day==7,"Weekend","Weekday"))))

steps_interval <- new_data %>% group_by(dayType,interval) %>% summarise(Average_steps = mean(New_steps,na.rm = T))

ggplot(steps_interval,aes(x=interval,y=Average_steps))+
    geom_line()+
    facet_grid(dayType~.)    
```

It can be seen that on weekends people walk fewer steps on average
